---
title: "Qfish Assignment"
author: "Jack Ochtman"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MB5370 Module 4 QFish

### Jack Ochtman

This script analyses shark drumline catch data from qfish.fisheries.qld.gov.au, separated into sharks \<2m in size and ≥2m in size. By plotting both size classes trends in the relative catch of small and large species/individuals can be observed, which may represent shark population trends.

## Initial Setup

Loads packages and dataset for analysis, sets and cleans column names.

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
# Install and load tidyverse and janitor packages, run install lines only if needed.
#install.packages("tidyverse")
library("tidyverse")
#install.packages("janitor")
library("janitor")
```

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
shark <- read_csv("../data/shark_size.csv") |>
  row_to_names(row_number = 1, remove_rows_above = FALSE) |>
  clean_names() # Load CSV and do basic formatting.  Open to validate.
shark
```

## Data wrangling

The dataset from QFish is unsuitable for analysis as-is. To suit this project financial year was converted to year, total values were removed, and size values were set to numeric (previously categorical).

```{r echo = T, results = 'hide'}
shark_size <- 
  shark |>
  rename(year = financial_year, small = x2m, large=x2m_2) |> # Corrects column names
  mutate(year = as.numeric(sub(".*-(\\d{2})", "20\\1", year))) |> # Changes FY to year, e.g. 2017-18 -> 2018
  filter(!is.na(year)) |> # Removes NA year values, this removes all "total" rows.
  mutate_at(vars(small, large), as.numeric) # Sets size data as numeric.
shark_size
```

The locations Bribie Island and Sunchine Coast contained incomplete data due to drumlines being installed later than in other locations. These locations, as well as their combined data, were removed prior to analysis. Additionally, the dataset had 0 values as blank cells, which were imported as "NA" and need to be changed to "0".

```{r echo = T, results = 'hide'}
shark_size <- shark_size %>%
  filter(area != "Bribie Island") %>%
  filter(area != "Sunshine Coast South & Bribie Island") %>%
  filter(area != "Sunshine Coast South") %>% # Remove locations with incomplete data
  mutate(across(where(is.numeric), ~replace_na(., 0))) # Replaces NA values with 0
shark_size
```

## Plotting

Following data wrangling results are plotted. Included are two unused plots showing \<2m and ≥2m sharks separately, these were useful for viewing trends. The finalised plot is faceted by location, allowing size differences to be seen more clearly.

```{r echo = T, results = 'show'}
ggplot(shark_size, aes(x = year, y = small, color = area)) +
  geom_line() +
  geom_point() + # Plots <2m shark catch as a line plot with points for each year.
  theme_minimal() +
    theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()) + # Remove grid
    labs(title = "Catch of <2m Sharks by Location",
  x = "Year",
  y = "No. of <2m Sharks Caught")
```

```{r echo = T, results = 'show'}
ggplot(shark_size, aes(x = year, y = large, color = area)) +
  geom_line() +
  geom_point() + # Plots ≥2m shark catch as a line plot with points for each year.
  theme_minimal() +
    theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()) + # Remove grid
    labs(title = "Catch of ≥m Sharks by Location",
  x = "Year",
  y = "No. of ≥2m Sharks Caught")
```

Below is the finalised facet plot, showing small and large size catch numbers for each location.

```{r echo = T, results = 'show'}
facet_plot <- ggplot() +
  geom_line(data = shark_size, aes(x = year, y = small, color = "<2m"), linewidth = 0.25) +
  geom_line(data = shark_size, aes(x = year, y = large, color = "≥2m"), linewidth = 0.25) +
  geom_point(data = shark_size, aes(x = year, y = small, color = "<2m"), shape = 16, size = 1) +
  geom_point(data = shark_size, aes(x = year, y = large, color = "≥2m"), shape = 17, size = 1) +
  facet_wrap(~ area) + # Seperate colours and point symbols for shark sizes, faceted by area.
  scale_color_manual(values = c("<2m" = "#E69F00", "≥2m" = "#56B4E9")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    strip.background = element_rect(fill = "white"),
    axis.text.x = element_text(angle = 90, hjust = 1)) + # Removes grid and sets background to white.
  labs(
    title = "Catch of Sharks by Size and Location",
    x = "Year",
    y = "No. of Sharks Caught",
    color = "Shark Size")

ggsave("../plots/shark_catch_plot.png", plot = facet_plot, width = 8, height = 6, dpi = 300) # Saves plot for publication

print(facet_plot) # Opens plot in markdown
```
